Reasoning About a Machine with Local Capabilities

### 2018-05-07 toplas/paper.tex
Minor changes to (mainly) section 5 of the ToPLaS paper.

  Our semantic model is not specific to a particular implementation of malloc,
  so rather than providing the implementation, we provide a specification that
  malloc must satisfy. The specification is a bit technical, but really just
  expresses standard expectations of how malloc should behave and this section
  can safely be skipped on first read.
  The malloc specification is presented in \figurename~\ref{fig:malloc-spec},
  and in the following, we will connect the description of the specification by
  referring to the item described.

@@

In addition to these syntactic requirements, we also specify standard intuitive
expectations for how malloc behaves. When malloc is invoked with a length
argument, then it will return a capability for a fresh piece of memory of that
size. It should be fresh in the sense that malloc has not already given out a
capability for any part of that memory before and will not do so in the future.
Finally, we also need to make sure that malloc behaves correctly when given a
nonsensical length argument such as a negative integer or a capability. However,
formulating these requirements is harder than one might expect. The problem is
that we want to allow malloc implementations to rely on internal state and
invariants on that state and it is reasonable to expect both this internal state
and the invariants to change over time. To express that malloc should behave
according to its specifications, but only if its internal state satisfies its
current invariants in an executing system, we will use the semantic model
defined in Section~\ref{sec:logical-relation} to express our specification.

@@

The malloc specification then dictates what malloc should do when invoked in a
memory with its internal state valid according to the malloc region (some future
evolution of the initial malloc region). Malloc is invoked with a length
argument. If this argument is invalid, that is, a negative integer or a
capability, we simply require malloc to fail (\ref{item:malloc-spec:fail}). This part of the specification
does not actually rely on the malloc region: for simplicity, we assume malloc
does not need its internal state to check the argument. When malloc is invoked
with a valid length (\ref{item:malloc-spec:correct-exec}), then it should return a fresh memory segment of the correct
length. This segment is required to come from the footprint of malloc, i.e.\ the
memory owned by the malloc region before the call. After malloc returns, we
require the malloc region to have evolved (according to the public future world
relation) to a new state where the new footprint is disjoint from the allocated
memory. This implies that future invocations of malloc can never return
previously-allocated memory.

@@

  \caption{The specification of malloc.}
  \label{fig:malloc-spec}
\end{figure}
As described previously, the specification of malloc ensures that malloc has no
capabilities pointing \emph{out} of malloc. It does, however, not say anything about
capabilities that point \emph{in} to malloc. We obviously can not
allow this if we want to trust malloc. We have chosen to keep the malloc
specification simple and let this assumption be in the lemmas that use malloc.
It is sufficient for these lemmas to require that there are no outside
capabilities for malloc in the initial configuration as capabilities cannot
appear out of thin air and the malloc specification makes sure that capabilities
are not leaked.

@@

The focus of this work is not to give a definitive specification and
implementation of malloc which is why we do not give a malloc implementation. We
are, however, confident that it is possible to make an implementation of malloc
that satisfies the malloc specification in Definition~\ref{spec:malloc}. There
are in particular two simplifications in our system that makes things easy:
First, we do not consider deallocation of memory which means that the data
structure malloc uses to keep track of free memory does not have to handle
reclaimed memory. Second, the malloc specification does not permit malloc to run
out of memory and thus refuse allocation. This is possible on our simple
capability machine because it has an infinite address space. An initial
capability with an infinite range of authority would of course need to be part
of malloc, but it could also double as the data structure that keeps track of
free memory. \itoplassug{If we have references for malloc being difficult, we
  could add that to the text.}
%\itoplassug{Add the assumption that the malloc capability in examples
%  satisfies specification.}

\section{Reusable macro instructions}
\label{sec:macros}
%\itoplassug{Add examples of macros to this section to make it clear what we mean by macros.\\}
%\itoplassug{Reviewer C, ESOP, asks for high-level specifications for the macros in the style of malloc. I am not sure we want to actually include this in the submission, so it is maybe not here we should put our attention. (see review in tex comment)\\}
% Finally, as already mentioned above, the approach relies on `call` and `scall` macro-instructions that are only outlined in the submission, and described by very complicated pseudocode and English prose in the companion technical appendix.  I still miss a fully formal description of 1- what instructions the macros expand to, exactly, and 2- what contract they satisfy (in the style of the contract for malloc in section 2 of the technical appendix).
With the semantic model defined, we would like to show its usefulness by proving
the correctness of a series of examples that rely
on well-bracketedness and local-state encapsulation and use the calling
convention to enforce these properties. However, the programs that
run on our capability are assembly programs and even program examples that would
be small in a high-level language become big and unintelligible at this
low level. Thus to make our program examples intelligible, we introduce a
series of low-level abstractions in the form of macros.
We define a number of reusable macros capturing the calling convention as well
as other conveniences. The macros that utilise the stack assume that it is
available in register $r_\stk$.
\itoplassug{Lau, suggestion: should we mention temporary registers? They are
  used in the macros, but it seems like a minor technical detail.}

The macro \texttt{\footnotesize{scall
    $r$($\overline{r_{\var{args}}}$,$\overline{r_{\var{priv}}}$)}} captures
those parts of the calling convention related to actually transferring control
to some adversarial code. Specifically, it pushes the contents of the private
registers, $\overline{r_{\var{priv}}}$, to the stack. Then it pushes the
``restoration'' code to the stack. The restoration code will be executed as the
first thing upon return, and restores the stack pointer and the old program
pointer. After the restoration code is pushed to the stack, a copy of the pc is
pushed to the stack after it has been adjusted to point to the first instruction
after the jump. From the stack pointer, a protected return pointer is created by
adjusting it to point to the return code and encapsulating it by restricting its
permission to Enter. Next, the stack pointer is restricted to the unused part,
and the unused part of the stack is cleared, so it does not govern any part of
the private stack frame which means that it can safely be given away. Finally,
the non-argument registers are cleared and we jump to $r$. Upon return after the
restoration code has been executed, the restoration code is popped from the
stack and the private words we pushed to the stack before the call are popped to
the private registers. The implementation of \texttt{scall} is presented in
\figurename~\ref{fig:scall-impl}, and the restoration code is presented in
\figurename~\ref{fig:scall-rest-code}. The implementation of \texttt{scall} uses
some of the macros we now present.

@@

The macros \texttt{\footnotesize{prepstack $r$}} and \texttt{\footnotesize{reqglob
    $r$}} are the last macros related to the calling convention. The former,
\texttt{prepstack} should be used when one receives a stack from an unknown
source. The \texttt{prepstack} macro first ensures the stack has permission
read/write-local/execute, and then it adjusts the stack capability, so it
follows the convention for the stack\footnote{The stack capability should always
point to the top most word on the stack. A stack received from an unknown source
can be treated as empty, so the stack capability should point just outside its
range of authority.}. The other macro, \texttt{reqglob}, ensures that the
capability in register $r$ is global. This macro should be used on callbacks
received from unknown sources to ensure that they cannot be derived from the
stack pointer.

The remaining macros are not directly related to the calling convention, but
they help making the examples in \sectionname~\ref{sec:examples} intelligible.
The macros \texttt{\footnotesize{push $r$}} and \texttt{\footnotesize{pop $r$}}
add and remove elements from the stack. The macro \texttt{\footnotesize{fetch
    $r$ $\var{name}$}} fetches the capability related to $\var{name}$ from the
linking table and stores it in register $r$. The macro
\texttt{\footnotesize{malloc $r$ $n$}} invokes malloc with size argument $n$.
The \texttt{malloc} macro assumes that a capability for malloc resides in the
linking table and is basically a fetch of the malloc capability followed by a
setup of a return pointer. Finally, the macro \texttt{\footnotesize{crtcls
    $\overline{(x_i,r_i)}$ $r$}} allocates a closure where $r$ points to the
closure's code and a new environment is allocated (using malloc) and the
contents of $\overline{r_i}$ is stored in the environment. In the code referred
to by $r$, an implicit fetch happens when an instruction refers to $x_i$.

@@

% Anti-reduction lemma
Out capability machine only allows the final memory of an execution to be
observed, so naturally the correctness lemmas we prove in
\sectionname~\ref{sec:examples} are statements about the memory in the halted
configuration. In order to prove something about the final memory, we encode it
as a region in a world and show that the initial configuration is in the
$\observations$-relation w.r.t.\ that world. The $\observations$-relation says
that if the initial configuration halts successfully, then the memory in the
halting state must still respect a private future world of the initial world. It
is, however, bothersome and error prone to try to argue about the entire
execution in one go. Instead we want to reason modularly, in the sense that we
only want to reason about parts of the execution at a time. To this end, we
prove an anti-reduction lemma that essentially says that if we can show that
an initial configuration, $\Phi$, steps to a configuration $\Phi'$ and
that $\Phi'$ is in the $\observations$-relation, then $\Phi$ is also in the
observation relation.
\begin{lemma}[Anti-reduction for $\observations$]
\label{lem:anti-red-obs}
  \begin{align*}
    & \forall n, n',i,\reg, \reg', \ms, \ms', \ms_r, W, W' \ldotp\\
    & \quad n' \geq n - i \land W' \futurestr W \land \\
    & \quad (\forall \ms_f \ldotp (\reg,\ms \uplus \ms_r \uplus \ms_f) \step[i] (\reg',\ms' \uplus \ms_r \uplus \ms_f)) \land \\
    & \quad \npair[n']{(\reg',\ms')} \in \observations(W') \\
    & \qquad \Rightarrow \npair{(\reg,\ms \uplus \ms_r)} \in \observations(W)
  \end{align*}
\end{lemma}

% FTLR
The anti-reduction works well when we reason about the execution of known code,
because we can execute it. When we want to reason about unknown code, the
anti-reduction lemma does not apply because we do not know what instructions are
being executed. This is where the FTLR (Theorem~\ref{thm:ftlr}) comes into play.
As a reminder, the FTLR says that if a capability only has access to safe values
with respect to a world, then it is safe to use it for execution in the same
world. The unknown code we consider will be assumed to only have access to safe
values, which typically means that we assume it has access to a linking table
with a malloc capability and otherwise consists of instructions. This allows us
to use the FTLR seeing as Malloc is a safe value cf.
Lemma~\ref{lem:malloc-in-vr} and instructions are really just integers and they
are always safe values. Here, it is important to remember that the FTLR gives us
that the capability for the untrusted code is safe w.r.t. some world, so even
though we described the usual assumption on the unknown code's memory, it is the
semantic model of the memory, i.e.\ the world, that the FTLR considers. In order
to argue that a specific configuration is safe, we need to argue that the
configuration is safe with respect to the world. That is, we need to argue that
the memory satisfies the world and the register-file is in the
$\stdrr$-relation. For instance, in the case where the untrusted code assumes
control first, we will use the FTLR to show that the capability for the unknown
code can be used for execution. The unknown code will get access to some known,
trusted code through a capability in the initial register-file. Because the
capability is in the register-file, we will have to argue that it is safe which
we do using Lemma~\ref{lem:anti-red-obs}.

@@

  % Note: $k_{\mathit{act-rec}}$ is the (standard) size of an activation record (2
  % words for stack pointer and return pointer and some fixed number of words for
  % the standard return instructions).
\end{lemma}
The essence of the \texttt{scall} lemma is to show \texttt{scall} is executed in
a reasonable state (\ref{item:scall:memsat}-\ref{item:scall:stack}), it is safe
to actually transfer control to the callee (\ref{item:scall:hyp-callee}), and that
the return capability given to the callee is safe (\ref{item:scall:hyp-cont}),
which corresponds to showing that the remainder of the program is safe to execute.

>>>>

  Our semantic model is not specific to a particular implementation of malloc,
  so rather than providing the implementation, we provide a specification that
  malloc must satisfy. The specification expresses standard expectations of how
  malloc should behave, but making it realistic requires using some of the
  technical machinery from the logical relation. As such, this section is a bit
  technical and can safely be skipped on first read. The malloc specification is
  presented in \figurename~\ref{fig:malloc-spec}, and in the following, we will
  connect the description of the specification by referring to the item
  described.

@@

In addition to these syntactic requirements, we also specify standard
expectations for how malloc behaves. Intuitively, we require that when malloc is
invoked with a length argument, then it will return a capability for a fresh
piece of memory of that size. It should be fresh in the sense that malloc has
not already given out a capability for any part of that memory before and will
not do so in the future. Also, when invoked with a nonsensical length argument
such as a negative integer or a capability, malloc should simply fail. However,
formulating these requirements is harder than one might expect. The problem is
that a realistic implementation of malloc needs to rely on internal state that
changes after every invocation, and relies on invariants on that state. We can
only expect that malloc behaves according to its specification, if its internal
state satisfies the current state of the invariants in an executing system. To
express this, we will use the semantic model defined in
Section~\ref{sec:logical-relation}.

@@

The malloc specification then dictates what malloc should do when invoked in a
memory with its internal state valid according to the malloc region (some future
evolution of the initial malloc region). If malloc is invoked with an invalid
length argument (that is, a negative integer or a capability), then we simply
require malloc to fail (\ref{item:malloc-spec:fail}). This part of the
specification does not actually rely on the malloc region: for simplicity, we
assume malloc does not need its internal state to check the argument. When
malloc is invoked with a valid length (\ref{item:malloc-spec:correct-exec}),
then it should return a fresh memory segment of the correct length. This segment
is required to come from the footprint of malloc, i.e.\ the memory owned by the
malloc region before the call. After malloc returns, we require the malloc
region to have evolved (according to the public future world relation) to a new
state where the new footprint is disjoint from the allocated memory. This
implies that future invocations of malloc can never return previously-allocated
memory.

@@

  \caption{The specification of malloc.}
  \label{fig:malloc-spec}
\end{figure}
As described previously, the specification of malloc ensures that malloc has no
capabilities pointing \emph{out} of malloc. It does, however, not say anything about
capabilities that point \emph{in} to malloc. We obviously cannot
allow this if we want to trust malloc. We have chosen to keep the malloc
specification simple and let this assumption be in the lemmas that use malloc.
It is sufficient for these lemmas to require that there are no outside
capabilities for malloc in the initial configuration as capabilities cannot
appear out of thin air and the malloc specification makes sure that capabilities
are not leaked.

@@

The reason we allow trusted code and adversary to invoke malloc is just to make
our work more realistic, but we are otherwise not interested in its details. As
such, we do not give a malloc implementation. We are, however, confident that it
is possible to make an implementation of malloc that satisfies the malloc
specification in Definition~\ref{spec:malloc}. There are in fact two
simplifications in our system that makes things easier: First, we do not consider
deallocation of memory which means that the data structure malloc uses to keep
track of free memory does not have to handle reclaimed memory. Second, the
malloc specification does not permit malloc to run out of memory and thus refuse
allocation. This is possible on our simple capability machine because it has an
infinite address space. An initial capability with an infinite range of
authority would of course need to be part of malloc, but it could also double as
the data structure that keeps track of free memory. % \itoplassug{If we have
  % references for malloc being difficult, we could add that to the text.}
%\itoplassug{Add the assumption that the malloc capability in examples
%  satisfies specification.}

\section{Reusable macro instructions}
\label{sec:macros}
%\itoplassug{Add examples of macros to this section to make it clear what we mean by macros.\\}
%\itoplassug{Reviewer C, ESOP, asks for high-level specifications for the macros in the style of malloc. I am not sure we want to actually include this in the submission, so it is maybe not here we should put our attention. (see review in tex comment)\\}
% Finally, as already mentioned above, the approach relies on `call` and `scall` macro-instructions that are only outlined in the submission, and described by very complicated pseudocode and English prose in the companion technical appendix.  I still miss a fully formal description of 1- what instructions the macros expand to, exactly, and 2- what contract they satisfy (in the style of the contract for malloc in section 2 of the technical appendix).
With the calling convention and logical relation defined, we would like to show its usefulness by proving
the correctness of a series of examples that rely
on well-bracketedness and local-state encapsulation and use the calling
convention to enforce these properties. However, the programs that
run on our capability machine are assembly programs and even program examples that would
be small in a high-level language become big and unintelligible at this
low level. Thus to make our program examples intelligible, we introduce a
series of low-level abstractions in the form of macros.
We define a number of reusable macros capturing the calling convention as well
as other conveniences. The macros that utilise the stack assume that it is
available in register $r_\stk$.
\itoplassug{Lau, suggestion: should we mention temporary registers? They are
  used in the macros, but it seems like a minor technical detail.}

The macro \texttt{\footnotesize{scall
    $r$($\overline{r_{\var{args}}}$,$\overline{r_{\var{priv}}}$)}} captures
those parts of the calling convention related to actually transferring control
to some adversarial code. Specifically, it pushes the contents of the private
registers, $\overline{r_{\var{priv}}}$, to the stack. Then it pushes the
``restoration'' code to the stack. The restoration code will be executed as the
first thing upon return, and restores the stack pointer and the old program
counter. After the restoration code is pushed to the stack, a copy of the pc is
pushed to the stack after it has been adjusted to point to the first instruction
after the jump. From the stack pointer, a protected return pointer is created by
adjusting it to point to the return code and encapsulating it by restricting its
permission to $\entry$. Next, the stack pointer is restricted to the unused
part, and the unused part of the stack is cleared (as discussed in
Section~\ref{sec:stack-and-return-pointer}). Finally, the non-argument registers
are cleared and we jump to $r$. Upon return after the restoration code has been
executed, the restoration code is popped from the stack and the private words we
pushed to the stack before the call are popped to the private registers. The
implementation of \texttt{scall} is presented in
\figurename~\ref{fig:scall-impl}, and the restoration code is presented in
\figurename~\ref{fig:scall-rest-code}. The implementation of \texttt{scall} uses
some of the macros we present next.

@@

The macros \texttt{\footnotesize{prepstack $r$}} and \texttt{\footnotesize{reqglob
    $r$}} are the last macros related to the calling convention. The former,
\texttt{prepstack} should be used when one receives a stack from an unknown
source. The \texttt{prepstack} macro first ensures the stack has permission
read/write-local/execute, and then it adjusts the stack capability, so it
follows the convention for the stack\footnote{The stack capability should always
point to the topmost word on the stack. A stack received from an unknown source
can be treated as empty, so the stack capability should point just outside its
range of authority.}. The other macro, \texttt{reqglob}, ensures that the
capability in register $r$ is global. This macro should be used on callbacks
received from unknown sources to ensure that they cannot be derived from the
stack pointer.

The remaining macros are not directly related to the calling convention, but
they help making the examples in \sectionname~\ref{sec:examples} intelligible.
The macros \texttt{\footnotesize{push $r$}} and \texttt{\footnotesize{pop $r$}}
add and remove elements from the stack. The macro \texttt{\footnotesize{fetch
    $r$ $\var{name}$}} fetches the capability related to $\var{name}$ from the
linking table and stores it in register $r$. The macro
\texttt{\footnotesize{malloc $r$ $n$}} invokes malloc with size argument $n$.
The \texttt{malloc} macro assumes that a capability for malloc resides in the
linking table and is basically a fetch of the malloc capability followed by a
setup of a return pointer. Finally, the macro \texttt{\footnotesize{crtcls
    $\overline{(x_i,r_i)}$ $r$}} allocates a closure where $r$ points to the
closure's code and a new environment is allocated (using malloc) and the
contents of $\overline{r_i}$ is stored in the environment. In the code referred
to by $r$, an implicit load from the environment happens when an instruction
refers to $x_i$.

@@

% Anti-reduction lemma
Our capability machine only allows the final memory of an execution to be
observed, so naturally the correctness lemmas we prove in
\sectionname~\ref{sec:examples} are statements about the memory in the halted
configuration. In order to prove something about some part of the final memory,
we establish an invariant for it as a region in a world and show that the
initial configuration is in the $\observations$-relation w.r.t.\ that world. The
$\observations$-relation says that if the initial configuration halts
successfully, then the memory in the halting state must still respect a private
future world of the initial world. It is, however, bothersome and error prone to
try to argue about the entire execution in one go. Instead we want to reason
modularly, in the sense that we only want to reason about parts of the execution
at a time. To this end, we prove an anti-reduction lemma that essentially says
that if we can show that an initial configuration, $\Phi$, steps to a
configuration $\Phi'$ and that $\Phi'$ is in the $\observations$-relation, then
$\Phi$ is also in the observation relation.
\begin{lemma}[Anti-reduction for $\observations$]
\label{lem:anti-red-obs}
  \begin{align*}
    & \forall n, n',i,\reg, \reg', \ms, \ms', \ms_r, W, W' \ldotp\\
    & \quad n' \geq n - i \land W' \futurestr W \land \\
    & \quad (\forall \ms_f \ldotp (\reg,\ms \uplus \ms_r \uplus \ms_f) \step[i] (\reg',\ms' \uplus \ms_r \uplus \ms_f)) \land \\
    & \quad \npair[n']{(\reg',\ms')} \in \observations(W') \\
    & \qquad \Rightarrow \npair{(\reg,\ms \uplus \ms_r)} \in \observations(W)
  \end{align*}
\end{lemma}

% FTLR
The anti-reduction works well when we reason about the execution of known code,
because we can execute it. When we want to reason about unknown code, the
anti-reduction lemma does not apply because we do not know what instructions are
being executed. This is where the FTLR (Theorem~\ref{thm:ftlr}) comes into play.
As a reminder, the FTLR says that if a capability only has access to safe values
with respect to a world, then it is safe to use it for execution in the same
world. The unknown code we consider will be assumed to only have access to safe
values, which typically means that we assume it has access to a linking table
with a malloc capability and otherwise consists of instructions. This allows us
to use the FTLR seeing as malloc is a safe value (cf.
Lemma~\ref{lem:malloc-in-vr}) and instructions are really just integers and they
are always safe values. Here, it is important to remember that the FTLR gives us
that the capability for the untrusted code is safe w.r.t. some world, so even
though we described the usual assumption on the unknown code's memory, it is the
semantic model of the memory, i.e.\ the world, that the FTLR considers. In order
to argue that a specific configuration is safe, we need to argue that the
configuration is safe with respect to the world. That is, we need to argue that
the memory satisfies the world and the register-file is in the
$\stdrr$-relation. For instance, in the case where the untrusted code assumes
control first, we will use the FTLR to show that the capability for the unknown
code can be used for execution. The unknown code will get access to some known,
trusted code through a capability in the initial register-file. Because the
capability is in the register-file, we will have to argue that it is safe which
we do using Lemma~\ref{lem:anti-red-obs}.

@@

  % Note: $k_{\mathit{act-rec}}$ is the (standard) size of an activation record (2
  % words for stack pointer and return pointer and some fixed number of words for
  % the standard return instructions).
\end{lemma}
Roughly, the \texttt{scall} lemma states that an invocation of scall is safe if
\texttt{scall} is executed in a reasonable state
(\ref{item:scall:memsat}-\ref{item:scall:stack}), the callee is safe to execute
(\ref{item:scall:hyp-callee}), and returning to the code after the scall is safe
(\ref{item:scall:hyp-cont}).
